---
title: "Untitled"
output: html_document
date: "2024-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import function for SLU MVM Data files and RKT output function.
```{r, include=FALSE}
import_slu_mvm_data_excel <- function(filename, numeric_var = NA, bad_quality_na = TRUE, sheet = 2){ # function for importing excel files into R
  #numeric var: the first column with measured numeric variables
  replace_less_than_var <- function(values, bad_quality_na){ # function for replacing values at the detection limit with half values
    values_parsed <- values %>% as.character() %>% parse_number(locale = locale(decimal_mark = ","))

    which_intervals <- which(substring(values,1,1) == "[")
    which_less_than <- which(substring(values,1,1) == "<")
    if (length(which_less_than) > 0) {
      values_less_than_half <- values[which_less_than] %>% gsub(",", ".", .) %>% gsub("<","",.) %>% as.numeric()
      values_parsed[which_less_than] <- values_less_than_half/2}

    if (bad_quality_na == TRUE) {values_parsed[which_intervals] <- NA}
    else{
      values_intervals <- values[which_intervals] %>%
        gsub("\\[","", .) %>%
        gsub("\\]","", .) %>%
        gsub(",",".", .) %>%
        as.numeric()
      values_parsed[which_intervals] <- values_intervals}

    return(values_parsed)
  }

  if (is.na(numeric_var) == T) {stop("Supply the column index of first measured numeric variable")}
out <- suppressWarnings(read_excel(filename, sheet = sheet,guess_max = 50000))
out <-
  
  mutate_at(out, c(numeric_var:ncol(out)),
            replace_less_than_var,
            bad_quality_na = bad_quality_na)
out <-
  mutate_at(out, c(1:(numeric_var - 1)),
            (function(x) {
                          x %>%
                            as.character() %>%
                            parse_guess(locale = locale(decimal_mark = ","))
              })
           )

}

rkt_out <- function(rkt){
  tibble(p.value = ifelse(is.na(rkt$sl.corrected)==T, rkt$sl, rkt$sl.corrected), statistic = rkt$S, slope = rkt$B, tau=rkt$tau)
}

```


Load packages
```{r, include=FALSE}
library(mgcv)
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
library(GWmodel)
library(knitr)
library(rkt)
library(compositions)
#library(ape)
library(ggpubr)
library(factoextra)
library(plsdepot)
library(ggrepel)
```

## Prepare data

Read data from the SLS, define year as the year of observation unless the month of observation is January, then the observation is counted for the year before. Variables are log-transformed (except Alk/Acid pH, vattentemperatur).
```{r, include=FALSE}
lakes <- import_slu_mvm_data_excel("SLS2023.xlsx", numeric_var = 29)%>%select(-c(`Siktdjup med kikare (m)`,`Siktdjup utan kikare (m)`))
lakesRMÖ <- import_slu_mvm_data_excel("SLS_RMÖ2023.xlsx", numeric_var = 29)%>%select(-c(`Slamhalt (mg/l)`,`SO4 (mg/l)`, `Syrgashalt (mg/l O2)`, `Syrgasmättnad (%)`, `Tot-N_ps (µg/l N)`, `Färgtal (mg Pt/l)`, `NO2-N (µg/l N)`,`NO3-N (µg/l N)`, `Tot-P_F (µg/l P)` ))%>%filter(Provtagningsmedium=="Sjö")


lakes%>%mutate(year1=year(Provdatum), 
               month=month(Provdatum),
               year=case_when(month==1~year1-1,
                               TRUE~year1)) %>%
        filter(year!=2007)%>%
        clean_names()%>%
        mutate(
          si_mg_l=case_when(si_mg_l<=0.5~0.25,
                            TRUE~si_mg_l),
          cl_mg_l=case_when(cl_mg_l<=0.25~0.125,
                            TRUE~cl_mg_l), 
          zn_mg_l=case_when(zn_mg_l<=0.5~0.25,
                            TRUE~zn_mg_l),
          v_mg_l=case_when(v_mg_l<=0.03~0.015,
                           TRUE~v_mg_l),
          pb_mg_l=case_when(pb_mg_l<=0.02~0.01,
                            TRUE~pb_mg_l),
          mn_mg_l=case_when(mn_mg_l<=0.5~0.25,
                            TRUE~mn_mg_l),
          ni_mg_l=case_when(ni_mg_l<=0.05~0.025,
                            TRUE~ni_mg_l),
          fe_mg_l=case_when(fe_mg_l<=10~5,
                            TRUE~fe_mg_l),
          f_mg_l=case_when(f_mg_l<=0.05~0.025,
                           TRUE~f_mg_l),
          cu_mg_l=case_when(cu_mg_l<0.04~0.02,
                            TRUE~cu_mg_l),
          cr_mg_l=case_when(cr_mg_l<=0.05~0.025,
                            TRUE~cr_mg_l),
          co_mg_l=case_when(co_mg_l<=0.01~0.005,
                            TRUE~co_mg_l),
          cd_mg_l=case_when(cd_mg_l<=0.006~0.003,
                            TRUE~cd_mg_l),
          as_mg_l=case_when(as_mg_l<=0.003~0.0015,
                            TRUE~as_mg_l),
          k_mg_l=case_when(k_mg_l<=0.04~0.02,
                           TRUE~k_mg_l),
          al_mg_l=case_when(al_mg_l<5~2.5,
                            TRUE~al_mg_l),
          nh4_n_mg_l_n=case_when(nh4_n_mg_l_n<=4~2,
                               TRUE~nh4_n_mg_l_n),
          no2_no3_n_mg_l_n=case_when(no2_no3_n_mg_l_n<=3~1.5,
                                     TRUE~no2_no3_n_mg_l_n),
          toc_mg_l_c=case_when(toc_mg_l_c<=0.05~0.025,
                               TRUE~toc_mg_l_c),
          totn=case_when(tot_n_t_nb_mg_l_n<=50~25,
                               TRUE~tot_n_t_nb_mg_l_n),
          totp=case_when(tot_p_mg_l_p<=1~0.5,
                               TRUE~tot_p_mg_l_p),
          absf=case_when(abs_f_420_5cm<=0.01~0.005,
                               TRUE~abs_f_420_5cm),
          so4_ic_mg_l_s=case_when(so4_ic_mg_l_s<=0.16~0.08,
                                  TRUE~so4_ic_mg_l_s),
          so4_ic_mg_l_so4=case_when(so4_ic_mg_l_so4<=0.48~0.24,
                                    TRUE~so4_ic_mg_l_so4),
          so4_mg_l_s=case_when(so4_mg_l_s<=0.2~0.1,
                               TRUE~so4_mg_l_s),
          DIN=nh4_n_mg_l_n+no2_no3_n_mg_l_n,
           BC=ca_mg_l+mg_mg_l+na_mg_l+k_mg_l,
          so4_komb=case_when(!is.na(so4_ic_mg_l_s)~so4_ic_mg_l_s/((32.06)/2), 
                                  !is.na(so4_ic_mg_l_so4)~so4_ic_mg_l_so4/((32.06+4*16)/2),
                                  TRUE~so4_mg_l_s/(32.06/2)))->lakes1a

lakes1<-lakes1a%>%
  select(-c(abs_f_254_5cm, abs_f_365_5cm, abs_f_436_m,abs_of_420_5cm, as_f_mg_l,cd_f_mg_l, co_f_mg_l,cr_f_mg_l, cu_f_mg_l, ni_f_mg_l, pb_f_mg_l,  so4_mg_l_s,so4_ic_mg_l_s, so4_ic_mg_l_so4, sr_mg_l, turb_fnu_fnu, u_mg_l,v_f_mg_l, zn_f_mg_l))%>%
  mutate(across(c(ca_mg_l:v_mg_l,zn_mg_l, absf, so4_komb) , log10, .names="log_{col}"))%>%
  select(-c(log_p_h))%>%
  rename(log_so4=log_so4_komb)
```

Include only lakes that are part of the random selection of the SLS.
```{r, include=FALSE}
slump<-read_excel("Slumpade Omdrevssjöprover tom 2023.xlsx")%>%clean_names()%>%select(md_mvm_id)

lakes1<-lakes1%>%filter(md_mvm_id %in% slump$md_mvm_id)
```

Read climate data and compute overall means, initial values (2008-2013) and change values. Join lakes data.
The climate data file can be obtained by the author. 
```{r, include=FALSE}
klimat<-read_excel("Klimatdata_omdrev.xlsx")%>%
  clean_names()%>%
  rename(temp_mean=temperature_c_mean1991to2020,
         precip_mean=precipitation_mm_mean1991to2020)%>%
  select(mvmid, temp_mean, precip_mean)

klimat2<-read_excel("Klimat_årsdata.xlsx")%>%clean_names()

klimat2 %>%
  filter(year>2007 & year<2014 & mvmid !=0)%>%
  group_by(mvmid) %>%
  summarize(across(c(temperature_c, precipitation_mm), mean, .names="{col}_start") %>%
  rename(temp_start=temperature_c_start, precip_start=precipitation_mm_start)) -> klimat_start

klimat2%>%
  filter(year>2011 & mvmid !=0)%>%
  nest(data=c(-mvmid, -smhi_id, -name, -east,-north, -station_id))%>%
  mutate(fit = map(data, possibly(~rkt(date = as.data.frame(.x)$year, y=as.data.frame(.x)$temperature_c, correct = TRUE), otherwise=NA_real_)),
         tidyfit = map(fit, possibly(~rkt_out(.x), otherwise=NA_real_))) %>% 
  unnest(tidyfit) %>%
  select(-data, -fit) %>%
  select(mvmid, slope) %>%
  unique()%>%
  rename(temp_change=slope) ->temp_change


klimat2 %>% 
  filter(year>2011 &mvmid !=0) %>% 
  nest(data=c(-mvmid, -smhi_id, -name, -east,-north, -station_id)) %>%
  mutate(fit = map(data, possibly(~rkt(date = as.data.frame(.x)$year, y=as.data.frame(.x)$precipitation_mm, correct = TRUE), otherwise=NA_real_)),
         tidyfit = map(fit, possibly(~rkt_out(.x), otherwise=NA_real_))) %>% 
  unnest(tidyfit) %>%
  select(-data, -fit) %>% 
  select(mvmid, slope) %>%
  unique() %>%
  rename(precip_change=slope) ->precip_change

lakes1<-lakes1 %>%
  filter(md_mvm_id %in% slump$md_mvm_id) %>%
  right_join(klimat, by=join_by(md_mvm_id==mvmid)) %>%
  right_join(temp_change, by=join_by(md_mvm_id==mvmid)) %>%
  right_join(precip_change, by=join_by(md_mvm_id==mvmid)) %>%
  right_join(klimat_start, by=join_by(md_mvm_id==mvmid))
```


Read GIS data and transform (package composition). Join lakes data.

```{r}
GISdata_landuse<-read_excel("all_outputs_omdrevssjöar_vivan2_zero_v3.xlsx", sheet="nmd_plc8")%>%clean_names()%>%select(-hav)%>%
  mutate(across(.cols=c("blandskog":"oppen_vatmark" ), .fns= ~ .x / 100))

GISdata<-read_excel("all_outputs_omdrevssjöar_vivan2_zero_v3.xlsx", sheet="aro_omdrevsjöar_240522")%>%clean_names()%>%select(mvm_stations_id, elev_utl)%>%rename(mvmid=mvm_stations_id)%>%full_join(GISdata_landuse)

as.data.frame(clr(GISdata%>%select(blandskog:oppen_vatmark)))%>%cbind(GISdata%>%select(mvmid, area_aro_km2,elev_utl))%>%filter(mvmid!=0)->GIS_trans

lakes1%>%full_join(GIS_trans, by=join_by("md_mvm_id"=="mvmid"))%>%filter(!is.na(overvakningsstation) & !is.na(stationskoordinat_n_x))->lakes1b
```

Set unique IDs for all lakes. If present the national station-ID is used, otherwise the MVM-ID.
```{r, include=FALSE}
#Identify which national stations ID are present
lakes1b %>%
  dplyr::select(stationskoordinat_e_y, stationskoordinat_n_x, nationellt_overvakningsstations_id) %>%
  group_by(stationskoordinat_n_x, stationskoordinat_e_y) %>%
  distinct() %>%
  slice(1) %>%
  filter(!is.na(nationellt_overvakningsstations_id)) %>%
  rename(new_ovv=nationellt_overvakningsstations_id)->stations1

#Join with original dataset to get the same ID for same coordinates
lakes1b%>%select(-nationellt_overvakningsstations_id) %>%
  left_join(stations1)%>%
  mutate(ID=new_ovv)->lakes_withID

#Identify mvm-ID for remaining lakes 
lakes_withID %>%
  dplyr::select(md_mvm_id) %>%
  group_by(md_mvm_id) %>%
  distinct() %>%
  slice(1) %>%
  filter(!is.na(md_mvm_id)) %>%
  mutate(new_ovv=md_mvm_id)->stations2

#Join with dataset to fill ID for stations that do not have a national ID
lakes_withID%>%
  select(-new_ovv) %>%
  left_join(stations2)%>%
  mutate(ID=case_when(is.na(ID)~as.character(new_ovv),
                      TRUE~ID))->lakes_a
```

Identify stations with only a single observations and remove them.

Remove stations with only one observation after 2011, since station-wise centering and, thus, trend analysis is not possible for them. Remove two stations with outlying values for Ca and Mg.
```{r, include=FALSE}
lakes_a %>%
  filter(year>2011) %>%
  drop_na(p_h) %>%
  group_by(ID, stationskoordinat_e_y, stationskoordinat_n_x) %>%
  mutate(n=n()) %>%
  filter(n<=1) %>%
  arrange(ID)%>%
  dplyr::select( stationskoordinat_e_y, stationskoordinat_n_x, overvakningsstation, ID)->only_single

lakes2 <- lakes_a %>%
  filter(!ID %in% only_single$ID) %>%
  filter(!ID %in%c("00202556", "00223105"))

```

Create initial variables all numeric variables 2008-2013

```{r}
lakes2 %>%
  filter(year<2014) %>%
  group_by(ID) %>%
  summarize(across(c(log_ca_mg_l:log_zn_mg_l,alk_acid_mekv_l, log_absf, p_h, log_so4), last, .names="{col}_start")) -> lakes_startvalue1

lakes2 %>%
  filter(year>=2014 & year< 2018) %>%
  group_by(ID)%>%
  summarize(across(c(log_ca_mg_l:log_zn_mg_l,alk_acid_mekv_l, log_absf, p_h, log_so4), first, .names="{col}_start2")) ->lakes_startvalue2

lakes_startvalue1 %>%
  full_join(lakes_startvalue2) %>%
  mutate(
    #log_al_mg_l_start=ifelse(is.na(log_al_mg_l_start), log_al_mg_l_start2, log_al_mg_l_start),
        # log_as_mg_l_start=ifelse(is.na(log_as_mg_l_start), log_as_mg_l_start2, log_as_mg_l_start),
         log_ca_mg_l_start=ifelse(is.na(log_ca_mg_l_start), log_ca_mg_l_start2, log_ca_mg_l_start),
         log_cd_mg_l_start=ifelse(is.na(log_cd_mg_l_start), log_cd_mg_l_start2, log_cd_mg_l_start),
         log_cl_mg_l_start=ifelse(is.na(log_cl_mg_l_start), log_cl_mg_l_start2, log_cl_mg_l_start),
         log_co_mg_l_start=ifelse(is.na(log_co_mg_l_start), log_co_mg_l_start2, log_co_mg_l_start),
         log_cr_mg_l_start=ifelse(is.na(log_cr_mg_l_start), log_cr_mg_l_start2, log_cr_mg_l_start),
         log_cu_mg_l_start=ifelse(is.na(log_cu_mg_l_start), log_cu_mg_l_start2, log_cu_mg_l_start),
         log_f_mg_l_start=ifelse(is.na(log_f_mg_l_start), log_f_mg_l_start2, log_f_mg_l_start),
         log_fe_mg_l_start=ifelse(is.na(log_fe_mg_l_start), log_fe_mg_l_start2, log_fe_mg_l_start),
         log_k_mg_l_start=ifelse(is.na(log_k_mg_l_start), log_k_mg_l_start2, log_k_mg_l_start),
         log_kond_25_m_s_m_start=ifelse(is.na(log_kond_25_m_s_m_start), log_kond_25_m_s_m_start2, log_kond_25_m_s_m_start),
         log_mg_mg_l_start=ifelse(is.na(log_mg_mg_l_start), log_mg_mg_l_start2, log_mg_mg_l_start),
         log_mn_mg_l_start=ifelse(is.na(log_mn_mg_l_start), log_mn_mg_l_start2, log_mn_mg_l_start),
         log_na_mg_l_start=ifelse(is.na(log_na_mg_l_start), log_na_mg_l_start2, log_na_mg_l_start),
         log_nh4_n_mg_l_n_start=ifelse(is.na(log_nh4_n_mg_l_n_start), log_nh4_n_mg_l_n_start2, log_nh4_n_mg_l_n_start),
         log_ni_mg_l_start =ifelse(is.na(log_ni_mg_l_start), log_ni_mg_l_start2, log_ni_mg_l_start),
         log_no2_no3_n_mg_l_n_start=ifelse(is.na(log_no2_no3_n_mg_l_n_start), log_no2_no3_n_mg_l_n_start2, log_no2_no3_n_mg_l_n_start),
         log_pb_mg_l_start=ifelse(is.na(log_pb_mg_l_start), log_pb_mg_l_start2, log_pb_mg_l_start),
         log_po4_p_mg_l_p_start=ifelse(is.na(log_po4_p_mg_l_p_start), log_po4_p_mg_l_p_start2, log_po4_p_mg_l_p_start),
         log_si_mg_l_start=ifelse(is.na(log_si_mg_l_start), log_si_mg_l_start2, log_si_mg_l_start),
         log_toc_mg_l_c_start=ifelse(is.na(log_toc_mg_l_c_start), log_toc_mg_l_c_start2, log_toc_mg_l_c_start),
         log_tot_n_t_nb_mg_l_n_start=ifelse(is.na(log_tot_n_t_nb_mg_l_n_start), log_tot_n_t_nb_mg_l_n_start2, log_tot_n_t_nb_mg_l_n_start),
         log_tot_p_mg_l_p_start=ifelse(is.na(log_tot_p_mg_l_p_start), log_tot_p_mg_l_p_start2, log_tot_p_mg_l_p_start),
         log_v_mg_l_start=ifelse(is.na(log_v_mg_l_start), log_v_mg_l_start2, log_v_mg_l_start),
         log_zn_mg_l_start=ifelse(is.na(log_zn_mg_l_start), log_zn_mg_l_start2, log_zn_mg_l_start),
         alk_acid_mekv_l_start=ifelse(is.na(alk_acid_mekv_l_start), alk_acid_mekv_l_start2, alk_acid_mekv_l_start),
         log_absf_start=ifelse(is.na(log_absf_start), log_absf_start2, log_absf_start),
         p_h_start=ifelse(is.na(p_h_start), p_h_start2, p_h_start),
         log_so4_start=ifelse(is.na(log_so4_start), log_so4_start2, log_so4_start))%>%
  select(ID, log_ca_mg_l_start:log_so4_start)->lakes_startvalue


```

Create variables describing the change between 2012 and 2023.
```{r, include=FALSE}
lakes2 %>%
  filter(year>2011) %>%
  group_by(ID) %>%
  arrange(year)%>%
  mutate(
    #Al_change=(last(log_al_mg_l)-first(log_al_mg_l))/(last(year)-first(year)),
    #     As_change=(last(log_as_mg_l)-first(log_as_mg_l))/(last(year)-first(year)),
         Ca_change=(last(log_ca_mg_l)-first(log_ca_mg_l))/(last(year)-first(year)),
         Cd_change=(last(log_cd_mg_l)-first(log_cd_mg_l))/(last(year)-first(year)),
         Cl_change=(last(log_cl_mg_l)-first(log_cl_mg_l))/(last(year)-first(year)),
         Co_change=(last(log_co_mg_l)-first(log_co_mg_l))/(last(year)-first(year)),
         Cr_change=(last(log_cr_mg_l)-first(log_cr_mg_l))/(last(year)-first(year)),
         Cu_change=(last(log_cu_mg_l)-first(log_cu_mg_l))/(last(year)-first(year)),
         F_change=(last(log_f_mg_l)-first(log_f_mg_l))/(last(year)-first(year)),
         Fe_change=(last(log_fe_mg_l)-first(log_fe_mg_l))/(last(year)-first(year)),
         K_change=(last(log_k_mg_l)-first(log_k_mg_l))/(last(year)-first(year)),
         Kond_change=(last(log_kond_25_m_s_m)-first(log_kond_25_m_s_m))/(last(year)-first(year)),
         Mg_change=(last(log_mg_mg_l)-first(log_mg_mg_l))/(last(year)-first(year)),
         Mn_change=(last(log_mn_mg_l)-first(log_mn_mg_l))/(last(year)-first(year)),
         Na_change=(last(log_na_mg_l)-first(log_na_mg_l))/(last(year)-first(year)),
         NH4_change=(last(log_nh4_n_mg_l_n)-first(log_nh4_n_mg_l_n))/(last(year)-first(year)),
         Ni_change=(last(log_ni_mg_l)-first(log_ni_mg_l))/(last(year)-first(year)),
         NO2NO3_change=(last(log_no2_no3_n_mg_l_n)-first(log_no2_no3_n_mg_l_n))/(last(year)-first(year)),
         Pb_change=(last(log_pb_mg_l)-first(log_pb_mg_l))/(last(year)-first(year)),
         PO4_change=(last(log_po4_p_mg_l_p)-first(log_po4_p_mg_l_p))/(last(year)-first(year)),
         Si_change=(last(log_si_mg_l)-first(log_si_mg_l))/(last(year)-first(year)),
         TOC_change=(last(log_toc_mg_l_c)-first(log_toc_mg_l_c))/(last(year)-first(year)),
         TOTN_change=(last(log_tot_n_t_nb_mg_l_n)-first(log_tot_n_t_nb_mg_l_n))/(last(year)-first(year)),
         TOTP_change=(last(log_tot_p_mg_l_p)-first(log_tot_p_mg_l_p))/(last(year)-first(year)),
         V_change=(last(log_v_mg_l)-first(log_v_mg_l))/(last(year)-first(year)),
         Zn_change=(last(log_zn_mg_l)-first(log_zn_mg_l))/(last(year)-first(year)),
         Alk_change=(last(alk_acid_mekv_l)-first(alk_acid_mekv_l))/(last(year)-first(year)),
         Absf_change=(last(log_absf)-first(log_absf))/(last(year)-first(year)),
         pH_change=(last(p_h)-first(p_h))/(last(year)-first(year)),
         SO4_change=(last(log_so4)-first(log_so4))/(last(year)-first(year)))%>%
          group_by(ID) %>%
  slice_tail(n=1) %>%
  ungroup %>%
  select(ID, ends_with("change"))->change_all
         
lakes2a<-lakes_startvalue%>%left_join(change_all)
  
lakes2b<-lakes2 %>%
  select(ID,area_aro_km2, blandskog:oppen_vatmark, precip_mean, temp_mean, area_aro_km2, elev_utl, temp_start, temp_change, precip_start, precip_change) %>%
  group_by(ID) %>%
  slice(1)
  

lakes2c<-lakes2 %>%
  select(ID, stationskoordinat_n_x, stationskoordinat_e_y) %>%
  unique()
```



```{r}
lakes2a %>%
  select(ID, ends_with("_start"), ends_with("change")) %>%
  left_join(lakes2b) %>%
  left_join(lakes2c) %>%
  drop_na() %>%
  rename(
    #Al_start=log_al_mg_l_start,
    #     As_start=log_as_mg_l_start,
         Ca_start=log_ca_mg_l_start,
         Cd_start=log_cd_mg_l_start,
         Cl_start=log_cl_mg_l_start,
         Co_start=log_co_mg_l_start,
         Cr_start=log_cr_mg_l_start,
         Cu_start=log_cu_mg_l_start,
         F_start=log_f_mg_l_start,
         Fe_start=log_fe_mg_l_start,
         K_start=log_k_mg_l_start,
         Kond_start=log_kond_25_m_s_m_start,
         Mg_start=log_mg_mg_l_start,
         Mn_start=log_mn_mg_l_start,
         Na_start=log_na_mg_l_start,
         NH4_start=log_nh4_n_mg_l_n_start,
         Ni_start=log_ni_mg_l_start, 
         NO2NO3_start=log_no2_no3_n_mg_l_n_start,
         Pb_start=log_pb_mg_l_start,
         PO4_start=log_po4_p_mg_l_p_start,
         Si_start=log_si_mg_l_start,
         TOC_start=log_toc_mg_l_c_start,
         TOTN_start=log_tot_n_t_nb_mg_l_n_start,
         TOTP_start=log_tot_p_mg_l_p_start,
         V_start=log_v_mg_l_start,
         Zn_start=log_zn_mg_l_start,
         Alk_start=alk_acid_mekv_l_start,
         Absf_start=log_absf_start,
         pH_start=p_h_start,
         SO4_start=log_so4_start,
         mixed_forest=blandskog,
         mixed_coniferous=gran_och_barrblandskog,
         vegetated_other=hedmark,
         artificial_sufaces=hardgjorda_ytor,
         other=ingen_tackning,
         deciduous_forest=lovskog,
         semi_urban=semiurbant,
         outcrop=hallmark,
         inland_water=sjo_och_vattendrag,
         forest_wetland=skog_pa_vatmark,
         pine_forest=tallskog,
         young_forest=ungskog,
         arable_land=akermark,
         open_wetland=oppen_vatmark,
         catchment_area=area_aro_km2,
         altitude=elev_utl) ->lakes_numeric

```

Run the PCA
```{r}
pH.pca <- prcomp(lakes_numeric[,c("Ca_start", "Cl_start", "Kond_start", "Mg_start", "Na_start",  "Alk_start", "pH_start", "SO4_start", "Ca_change", "Cl_change", "Kond_change", "Mg_change", "Na_change",   "SO4_change", "mixed_forest",
         "mixed_coniferous", "vegetated_other", "artificial_sufaces", "other", "deciduous_forest",       "semi_urban", "inland_water", "forest_wetland", "pine_forest", "young_forest", "arable_land",
         "open_wetland", "outcrop","temp_mean", "precip_mean", "temp_start", "precip_start", "temp_change", "precip_change", "catchment_area",  "altitude")], center=TRUE,scale=TRUE)  
```

Visualize the PCA
```{r}

fviz_pca_var(pH.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE ,  # Avoid text overlapping
             title=""  
             )

fviz_pca_var(pH.pca,
             axes=c(3,4),
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```


```{r}
# # Eigenvalues
# eig.val <- get_eigenvalue(pH.pca)
# eig.val
  
# # Results for Variables
# res.var <- get_pca_var(pH.pca)
# res.var$coord          # Coordinates
# res.var$contrib        # Contributions to the PCs
# res.var$cos2           # Quality of representation 
# 
# # Results for individuals
# res.ind <- get_pca_ind(pH.pca)
# res.ind$coord          # Coordinates
# res.ind$contrib        # Contributions to the PCs
# res.ind$cos2           # Quality of representation 
# 
# 
# qqnorm(res.ind$coord)  #check for outliers?
```


The station-wise mean-centered variable is computed for all analysed variables to remove site_specific variation. 
```{r, include=FALSE}

lakes2 %>%
  filter(year>2011) %>%
  group_by(ID, stationskoordinat_e_y, stationskoordinat_n_x) %>%
  summarize(mean_ph=mean(p_h, na.rm=TRUE))->lakes_mean

lakes2 %>%
  filter(year>2011) %>%
  left_join(lakes_mean) %>%
  mutate(pH_cent=p_h - mean_ph,year)->lakes3
```


Extract the coordinates for individual observations from the PCA abd combine with the original lakes data set.
```{r}
res.ind <- get_pca_ind(pH.pca)


lakes_numeric %>%
  select(ID) %>%
  bind_cols(res.ind$coord) %>%
  right_join(lakes3) %>%
  drop_na(Dim.1, pH_cent)->lakes3_pca_res
```

Thematically weighted regression models on PCA coordinates for years 2008 to 2023

```{r}
lakes3_pca_res%>%
  filter(year>2011)->lakes3_pca_res2011

lakes.spdf <- SpatialPointsDataFrame(lakes3_pca_res2011[, c("Dim.1", "Dim.2")], lakes3_pca_res2011)

distances<-gw.dist(coordinates(lakes.spdf))

bw<-bw.gwr(pH_cent ~ year,adaptive=TRUE,kernel="bisquare", parallel.method="omp",data=lakes.spdf)

model_pH_pca<-gwr.basic(pH_cent ~ year, adaptive=TRUE, bw=bw, kernel="bisquare", parallel.method="omp",data=lakes.spdf)


hist(model_pH_pca$SDF$Local_R2)
summary(model_pH_pca$SDF$Local_R2)
summary(model_pH_pca$SDF$year)
```

Figure : Thematically weighted regression coefficients on PCA coordinates
```{r}
lakes3_pca_res2011$trend<-model_pH_pca$SDF$year

lakes3_pca_res2011%>%arrange(abs(trend))%>%ggplot( aes(x=Dim.1,y=Dim.2))+
  geom_point(aes(colour=trend))+
  scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(colour="Linear trend slope")+
  theme_classic()+
  ylab("PCA component 2")+
  xlab("PCA component 1")+
 theme(axis.title = element_text(size=16), legend.key.size = unit(1, 'cm'))
```

Thematically weigthed regression model on the third and fourth PCA componentes for supplementary,
```{r}
lakes3_pca_res%>%filter(year>2011)->lakes3_pca_res2011

lakes34.spdf <- SpatialPointsDataFrame(lakes3_pca_res2011[, c("Dim.3", "Dim.4")], lakes3_pca_res2011)
distances34<-gw.dist(coordinates(lakes34.spdf))

bw34<-bw.gwr(pH_cent ~ year,adaptive=TRUE,kernel="bisquare", parallel.method="omp",data=lakes34.spdf)

model_pH_pca34<-gwr.basic(pH_cent ~ year, adaptive=TRUE, bw=bw34, kernel="bisquare", parallel.method="omp",data=lakes34.spdf)

lakes3_pca_res2011$trend<-model_pH_pca34$SDF$year
lakes3_pca_res2011%>%arrange(abs(trend))%>%ggplot( aes(x=Dim.3,y=Dim.4))+
geom_point(aes(colour=trend))+
 
scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(colour="Linear trend slope")+
  theme_classic()+
  ylab("PCA component 4")+
  xlab("PCA component 3")+
     theme(axis.title = element_text(size=16), legend.key.size = unit(1, 'cm'))
```

Fit PLS 
```{r}
pls_pH = plsreg1(lakes_numeric[,c("Ca_start", "Cl_start", "Kond_start", "Mg_start", "Na_start",  "Alk_start", "pH_start", "SO4_start", "Ca_change", "Cl_change", "Kond_change", "Mg_change", "Na_change",  "SO4_change", "mixed_forest",
         "mixed_coniferous", "vegetated_other", "artificial_sufaces", "other", "deciduous_forest",       "semi_urban", "inland_water", "forest_wetland", "pine_forest", "young_forest", "arable_land",
         "open_wetland", "outcrop","temp_mean", "precip_mean", "temp_start", "temp_change", "precip_start", "precip_change", "catchment_area", "altitude")], lakes_numeric[, "pH_change", drop = FALSE], comps = 12)
```

Use updated function for plotting PLS results
```{r}
source("plot.plsreg1 with ggplot_changed.R")
rbPal <- colorRampPalette(c("#00AFBB", "#E7B800", "#FC4E07"))


df<-cbind(pls_pH$R2Xy[1:36,1], pls_pH$R2Xy[1:36,2])
max <- df[,2] #df is cumulative, extract the R2 for both components together
plot_pls<-plot_plsreg1(pls_pH, comps = c(1, 2), where=c("t","t"), main="", 
             length=0.1, angle=30, lwd=0.8, col_xarrows = rbPal(50)[as.numeric(cut(max,breaks = 50))] ,
     col_xlabels = rbPal(50)[as.numeric(cut(max,breaks = 50))], col_yarrows = "black",
     col_ylabel="black")+
  theme(axis.title = element_text(size=14), legend.key.size = unit(0.6, 'cm'))



plot_colors<-ggplot(NULL, aes(x=1, y=seq(0,0.8,length.out=50), col=seq(0,0.8,length.out=50)))+
         geom_point()+
        scale_color_gradient2(low="#00AFBB", mid="#E7B800", high="#FC4E07", midpoint = 0.4)+
        labs(color="contrib")+
        theme(axis.title = element_text(size=16), legend.key.size = unit(0.8, 'cm'))

library(cowplot)
library(grid)
legend<-get_legend(plot_colors)

legend_plot<-as_ggplot(legend)

ggarrange(plot_pls, legend_plot,NULL,  ncol=3, widths = c(4, 0.7, 1))

```


Combine PLS coordinates with original lakes data
```{r}
lakes_numeric%>%select(ID)%>%bind_cols(pls_pH$x.scores[,1:2])%>%right_join(lakes3)%>%drop_na(t1, pH_cent)->lakes3_pls_res

 lakes3_pls_res%>%filter(year>2011)->lakes3_pls_res2011

lakes.spdf <- SpatialPointsDataFrame(lakes3_pls_res2011[, c("t1", "t2")], lakes3_pls_res2011)
distances<-gw.dist(coordinates(lakes.spdf))

bw<-bw.gwr(pH_cent ~ year,adaptive=TRUE,kernel="bisquare", parallel.method="omp",data=lakes.spdf)

model_pH_pls<-gwr.basic(pH_cent ~ year, adaptive=TRUE, bw=bw, kernel="bisquare", parallel.method="omp",data=lakes.spdf)

summary(model_pH_pls$SDF$year)
pval_pH<-gwr.t.adjust((model_pH_pls))


hist(model_pH_pls$SDF$Local_R2)
```

```{r}
lakes3_pls_res2011$trend<-model_pH_pls$SDF$year

lakes3_pls_res2011%>%arrange(abs(trend))%>%ggplot(aes(x=t1,y=t2))+
 geom_point(aes(colour=trend))+
 scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(colour="Linear trend slope")+
  theme_classic()+
  ylab("PLS component 2")+
  xlab("PLS component 1")+
   theme(axis.title = element_text(size=16), legend.key.size = unit(1, 'cm'))
```


Read coordinates that define the border of Sweden, including the largest lakes. 
```{r}
border <- read_excel("sverige_smooth_coordinates_SWEREF.xlsx")
border1<-subset(border, namn=="sverige", select=c(East_SWERE, North_SWER))
border_vänern<-subset(border, namn=="vänern", select=c(East_SWERE, North_SWER))
border_vättern<-subset(border, namn=="vättern", select=c(East_SWERE, North_SWER))
border_mälaren<-subset(border, namn=="mälaren", select=c(East_SWERE, North_SWER))

```

Figure 5 (right): TWR-PCA in a geographical perspective
```{r}
lakes3_pca_res2011%>%cbind(model_pH_pca$SDF$year)->lakes3_pca_res2011_a
gwr.point_geographical_pca<-lakes3_pca_res2011_a%>% arrange(abs(`model_pH_pca$SDF$year`))%>%
  ggplot(aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=`model_pH_pca$SDF$year`), size=1)+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
 scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(color="Linear trend slope")+
  theme_classic()+
  theme(legend.position="bottom")+
  ylab("")+
  xlab("")
```

Figure 5 (mid): TWR-PLS in a geographical perspective
```{r}
lakes3_pls_res2011%>%cbind(model_pH_pls$SDF$year)->lakes3_pls_res2011_a

gwr.point_geographical<-lakes3_pls_res2011_a%>% arrange(abs(`model_pH_pls$SDF$year`))%>%
  ggplot(aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=`model_pH_pls$SDF$year`), size=1)+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(color="Linear trend slope")+
  theme_classic()+
  theme(legend.position="bottom")+
  ylab("")+
  xlab("")


```

Figure 5 (left): Geographically weighted regression model
```{r}
lakes_after_2011<-lakes3%>%filter(year>2011)%>%drop_na(pH_cent)
lakes_after.spdf <- SpatialPointsDataFrame(lakes_after_2011[, c(5,4)], lakes_after_2011)

bw_after<-bw.gwr(pH_cent ~ year,adaptive=TRUE,kernel="bisquare", parallel.method="omp",data=lakes_after.spdf)

model_pH<-gwr.basic(pH_cent ~ year,bw=bw_after,adaptive=TRUE,kernel="bisquare", parallel.method="omp",data=lakes_after.spdf)

lakes_after_2011%>%cbind(model_pH$SDF$year)->lakes_after_2011_a

gwr.point_pH<-lakes_after_2011_a%>%arrange(abs(`model_pH$SDF$year`))%>%
  ggplot( aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=`model_pH$SDF$year`), size=1)+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(color="Linear trend slope")+
  theme_classic()+
  theme(legend.position="bottom")+
  ylab("")+
  xlab("")



```

Combine TWR-PCA, TWR-PLS and GWR into one plot
```{r}
library(ggpubr)
ggarrange(gwr.point_geographical_pca, gwr.point_geographical, gwr.point_pH, ncol=3, common.legend = TRUE, legend="bottom")

```

```{r}


plot1<- lakes3_pls_res2011%>%arrange(abs(trend))%>%ggplot(aes(x=t1,y=t2, color=trend, text=ID))+
 geom_point()+
 scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(colour="Linear trend slope")+
  theme_classic()+
  ylab("PLS component 2")+
  xlab("PLS component 1")+
   theme(axis.title = element_text(size=16), legend.key.size = unit(1, 'cm'))
```



```{r}
# lakes_numeric_without_extremes<-lakes_numeric%>%filter(!ID %in% c("00263474",   "00202608", "00202639", "00203743", "00262280", "00263325", "00211399", "00184188", "00184201", "00185193", "00185616", "00202147", "00202288", "00202293", "00202402", "00202411", "00202414", "00202427", "00184186", "00185253", "00185555", "00185559", "00260644", "00260672", "00260843", "00261325", "00206076", "00243359", "00260925", "00256352", "39844" )) 


lakes_numeric_without_extremes<-lakes_numeric%>%filter(!ID %in% c("00211399", "00203743", "00256352", "00202608", "00256352", "39844", "00262280", "00202639", "00263325", "00260874", "00261161", "00263474", "00202414", "00263291", "00256501", "00261408", "00263381", "00260874")) 


pls_pH_without_extremes = plsreg1(lakes_numeric_without_extremes[,c("Ca_start", "Cl_start", "Kond_start", "Mg_start", "Na_start",  "Alk_start", "pH_start", "SO4_start", "Ca_change", "Cl_change", "Kond_change", "Mg_change", "Na_change",   "SO4_change", "mixed_forest",
         "mixed_coniferous", "vegetated_other", "artificial_sufaces", "other", "deciduous_forest",       "semi_urban", "inland_water", "forest_wetland", "pine_forest", "young_forest", "arable_land",
         "open_wetland", "outcrop","temp_mean", "precip_mean", "temp_start", "temp_change", "precip_start", "precip_change", "catchment_area",  "altitude")], lakes_numeric_without_extremes[, "pH_change", drop = FALSE], comps = 3)

rbPal <- colorRampPalette(c("#00AFBB", "#E7B800", "#FC4E07"))
rbPal(50)[as.numeric(cut(pls_pH$R2Xy[1:36,1],breaks = 50))]
df<-cbind(pls_pH$R2Xy[1:36,1], pls_pH$R2Xy[1:36,2])
max <- apply(df, 1, max)
plot_plsreg1(pls_pH_without_extremes, comps = c(1, 2), where=c("t","t"), main="", 
             length=0.1, angle=30, lwd=0.8, col_xarrows = rbPal(50)[as.numeric(cut(max,breaks = 50))] ,
     col_xlabels = rbPal(50)[as.numeric(cut(max,breaks = 50))], col_yarrows = "black",
     col_ylabel="black")
```

```{r}
lakes_numeric_without_extremes%>%select(ID)%>%bind_cols(pls_pH_without_extremes$x.scores[,1:2])%>%right_join(lakes3)%>%drop_na(t1, pH_cent)->lakes3_pls_res_without_extremes


lakes3_pls_res_without_extremes%>%filter(year>2011)->lakes3_pls_res2011_without_extremes

lakes.spdf <- SpatialPointsDataFrame(lakes3_pls_res2011_without_extremes[, c("t1", "t2")], lakes3_pls_res2011_without_extremes)
distances<-gw.dist(coordinates(lakes.spdf))

bw<-bw.gwr(pH_cent ~ year,adaptive=TRUE,kernel="bisquare", parallel.method="omp",data=lakes.spdf)

model_pH_pls_without_extremes<-gwr.basic(pH_cent ~ year, adaptive=TRUE, bw=bw, kernel="bisquare", parallel.method="omp",data=lakes.spdf)
```

```{r}
lakes3_pls_res2011_without_extremes$trend<-model_pH_pls_without_extremes$SDF$year

lakes3_pls_res2011_without_extremes%>%arrange(abs(trend))%>%ggplot(aes(x=t1,y=t2))+
   geom_point(aes(colour=trend))+
  #coord_fixed()+
  scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(colour="Linear trend slope")+
  theme_classic()+
  ylab("")+
  xlab("")
```

```{r}
lakes3_pls_res2011_without_extremes%>%cbind(model_pH_pls_without_extremes$SDF$year)->lakes3_without_extremes
gwr.point_pH<-lakes3_without_extremes%>%arrange(abs(`model_pH_pls_without_extremes$SDF$year`))%>%
  ggplot( aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=`model_pH_pls$SDF$year`), size=1)+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_gradientn(colors=c("darkblue", "lightblue3", "darkolivegreen1", "yellow", "orange", "brown2", "darkred"),values=c(0, 40, 45,  50, 55, 60, 100)/100,
                    na.value = "grey50", guide = "colourbar", limits=c(-0.1, 0.1))+
  labs(color="Linear trend slope")+
  theme_classic()+
  theme(legend.position="bottom")+
  ylab("")+
  xlab("")
```




